---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Literature Review

Interactive data visualization is becoming more and more popular. Interactive figures appear everywhere, from news articles, to business-analytic dashboards, to science communication outlets such as journal websites and personal blogs. Yet, despite their popularity, there seem to be a lot of unanswered questions about them.   

## What even is interactivity? {#what-is-interactivity}

> If it looks like a duck, swims like a duck, and quacks like a duck, then it probably is a duck.
> 
> [...] The irony is that while the phrase is often cited as proof of abductive reasoning, it is not proof, as the mechanical duck is still not a living duck
>
> [Duck Test](https://en.wikipedia.org/wiki/Duck_test) entry, [@wikipedia2022] 

What is an interactive visualization? It may seem like it should be easy to answer this question, given that interactive visualizations are so widely used. Unfortunately, that does not seem to be the case. When researchers and applied practitioners say "interactive visualization", they tend to refer to many different things across many different contexts.  

Firstly, it is necessary to disambiguate whether we refer to "interactive visualization" as a noun, a thing that exists, or to "interactive visualization" as a process, an action undertaken by a human being. Pike et al. [-@pike2009] note that "interaction" is an overloaded term that can refer to either the concrete set of tools through which users manipulate the visual information or to the more abstract "human interaction with information" - the back-and-forth between the user and the visual information presented to them [see also @yi2007toward]. Which definition is used depends largely on the field: statisticians and computer scientists tend to talk about interactive visualizations as things, whereas researchers in the fields of cognitive science and human-computer interaction tend to emphasize interactive visualization as action, although there is also considerable overlap. While the cognitive aspect of humans interacting with visual information is definitely interesting, for the purpose of this thesis, I will mainly discuss "interactive visualizations" as things, i.e. visual objects that live on a computer screen.

Yet, even if we narrow down the focus on interactive visualizations as visual objects, there is still a lot of ambiguity. Some researchers define interactive visualizations very broadly. For example, there are those that talk about interactive visualizations as any visualizations can be actively manipulated by the user [@brodbeck2009]. To others, the key thing about interactivity is time, or rather, the lag between the user's input and changes to the visualization, with less lag meaning more interactivity [@becker1987; @buja1996interactive]. Some even make the distinction between "interactive" and "dynamic" manipulation, where interactive manipulation happens discretely such as by pressing a button or selecting an item from a drop-down menu, whereas dynamic manipulation happens continuously, in real-time, for example by smoothly moving a slider or by clicking-and-dragging [@rheingans2002we; @jankun2007model]. What these two definitions - which I will label the "basic" and "temporal" - have in common is that they impose relatively few restrictions on what kinds of visualizations can be considered interactive. For example, one could argue if the user runs a code from a command line to generate a new plot, this could be considered interactive visualization, under these definitions.  

In contrast, other researchers consider the category of "interactive visualizations" to be much narrower. For many, the defining features are the ability to query different parts of the dataset (by e.g. zooming, panning, and filtering), and the propagation of changes between connected or "linked" parts of the visualization [@kehrer2012generic; @buja1996interactive; @keim2002information; @unwin1999mar]. Similarly, in Visual Analytics (VA) research, a distinction is made between "surface-level" (or "low-level") and "parametric" (or "high-level") interactions, where surface-level interactions manipulate attributes of the visual domain only (e.g. zooming and panning), whereas parametric interactions manipulate attributes of mathematical models or algorithms, underlying the visualization [@leman2013, @pike2009]. These definitions of interactivity - which I will label "querying", "linked", and "parametric", respectively - can be used to classify considerably more complex objects. Specifically, while some of the querying interactions like zooming and panning can be implemented by manipulating graphical attributes of the visualization only, linked and parametric interactions require special underlying data structures in all but the most simple cases. For example, in the case of the prototypical interactive feature called "linked selection", the user can click-and-drag to select points on a scatter plot and the selection will then be reflected in another plot, for example, by drawing a bars in a barplot with heights corresponding to the number of the selected points. This requires the barplot to be "aware" of which cases in the data (corresponding to the points in the scatterplot) have been selected, and it has to respond to these changes online, in real time, by recomputing the heights of the newly drawn bars. 

Put simply, the term "interactive visualization" means a lot of different things to a lot of different people. Below, in Table \@ref(tab:definitions), I have attempted to summarize the main types of interactivity of data visualizations. This list is not supposed to be exhaustive as more complete and detailed taxonomies of interactive visualizations have been described [see @yi2007toward]. Instead, the point of this list is to provide a rough sketch of the space interactive visualizations live in: 

```{r definitions, echo = FALSE}

tab <- read.csv('tables/definitions.csv')
colnames(tab) <- c("Type", "Short definition", "Details")
knitr::kable(tab, longtable = TRUE, caption = 'Types of interactivity')

```

Using the term "interactive" to cover such a wide range of visualizations necessarily leads to ambiguity. For example, imagine we have a single scatterplot. The user can click-and-drag on this scatterplot to highlight certain points (i.e. brush). Should this scatterplot be considered "interactive"? Under the most basic definition, yes, since the user *can* affect the visualization through interaction. However, some researcher may not be satisfied with that and may ask us to tell them whether the visualization meets linked or parametric defition of interactivity. With only a single plot, there's nowhere for our interactions to propagate to, and changing the color of individual points can hardly be considered a change to a mathematical model, so our verdict would then have to be "not interactive". What if the researcher was a fan of the temporal definition? If we assume the results of the user's interactions render smoothly enough, we could certainly call the plot interactive. However, what if more data is added to the plot and the user now has to wait several hundreds of milliseconds, or even seconds, before the interaction is rendered? Does an interactive visualization stop being interactive as a result of our computational resources being reduced? Therefore, even with such a simple example, there are many arguments that could be made for and against calling a plot an "interactive visualization".

```{r, message=FALSE, warning=FALSE}

library(plotly)
plot_ly(mtcars, x = ~wt, y = ~mpg)

```

>Is this an interactive visualization? Depends on who you ask.

It is possible to rank these types of interactivity on different metrics. The most obvious is perhaps the programming and computational complexity: how large codebase and how much computational resources is required implement this type of interactivity? Clearly, the basic or surface-level definition is the least restrictive, and so the simplest and the computationally most cheap interactive visualization systems will be basic. Refering back to the example of the "dubiously interactive" scatterplot, we can implement the highlighting by assigning each point in it a color attribute. Then, whenever the user click-and-drags, all we need to do is to manipulate that color attribute. We do not need to refer back to the data or do any additional computation, and this makes the interaction very cheap. Things get a bit more complicated with temporally interactive visualizations, as there is now the requirement that the visualizations respond fast enough to the user's interactions. This adds some computational complexity, especially when the data volume is high, however if the interactive action is simple then this can still be done very efficiently. Next, querying interactivity can be somewhat more complex again, since while zooming and panning can be implemented very cheaply, by manipulating only the graphical attributes of the visualization (i.e. axis limits) only, filtering requires an additional data structure (a variable) to keep track of which cases in the data are currently filtered. Finally, linked and parametric will usually be the most complex, both programmatically and computationally, as they typically require extra data structures and online computations. However, there is nuance. For example, if we want to implement linked plots that all show one-to-one representations of the cases in the data (e.g. points on a scatterplot), we might be able to get away with a single variable that tracks which cases are selected, similarly to filtering. However, if we also want to include plots with many-to-one representation of the data (e.g. barplot: the count of many cases describes the height of one bar), then we need to be able to compute the relevant summary statistics (height of the bar) on the fly. That is, in addition of having a variable which keeps track of which cases are selected, the plots also need to "remember" how their summary statistics are computed and be able to re-do the computation each time the selection status changes.        

```{r, echo=FALSE}

label <- c("Basic/Surface-level", "Temporal", 
           "Querying", "Linked", "Parametric")
comp <- c(1, 5, 3, 4, 5)
program <- c(1, 2, 3, 5, 5)

usr <- par("usr")


plot(0:6, 0:6, type = "n", 
     xlab = "", ylab = "", axes = FALSE)
text(program, comp, labels = label)
arrows(0, 0, c(0, 6), c(6, 0), length = 0.1, 
       angle = 20, xpd = TRUE)
title(xlab = "Program complexity", line = 1)
title(ylab = "Computational complexity", line = 1)

```

Another way to rank the types of interactivity is by their usefulness. I argue that this is inversely proportional to the programming and computational complexity. Parametric and linked interactions are the most useful since they allow the user to get a high-level understanding of the patterns and interactions within the data [@pike2009; @leman2013]. Being able to ask questions such as: "which bar on a barplot do these points on a scatterplot belong to?" is akin to the conditioning operator in probability, and as such extremely useful. Querying interactions are fairly useful too, since the allow the user to "drill down" and answer questions about specific parts and subsets of the dataset [@dix1998]. Temporal interactivity is useful in a "necessary but not sufficient" way, in that a large lag between the user's input and changes to the graphics renders any kind of interaction moot, but a visualization is not guaranteed to be useful, no matter how smoothly it renders. Finally, basic interaction may add little usefulness above and beyond the static plot that would exist without it. 

## Exploration vs. Presentation

## Brief History of Interactive Data Visualization

Static data visualization has a rich and an intricate history, and a full treatment would be beyond the scope of the current thesis [but see e.g. @dix1998; @friendly2006; @friendly2021]. Suffice it to say, prior to the second half of the 20th century, data visualization was largely seen as at best secondary to "serious" statistical analysis, although there were indeed some prominent examples [@friendly2006]. However, starting at the end of 1950's, a series of developments lead to a great increase in prominence. Firstly, at the theoretical level, the work of Tukey [-@tukey1962; -@tukey1977] and Bertin [-@bertin1967] established data visualization as an independent and valuable discipline in its own right. Secondly, at the practical level, the development of personal computers [see e.g. @abbate1999] and high-level programming languages, most notably FORTRAN in 1954 [@backus1978], introduced powerful and widely available tools that rendered the production of figures near-effortless, in comparison to the earlier hand-drawn techniques. Combined, these developments lead to a surge in the use and dissemination of data visualization.  

Interactive visualization would not be left far behind. Early systems such as that of Fowlkes [-@fowlkes1969] were typically very specialized and focused on a narrow range of tasks, such as displaying probability distributions under ranging parameters.      

The older generation interactive visualization systems seemed to have been designed with interesting types of interactivity in mind. For example, the popular XGobi [@swayne1998] and GGobi [@swayne2003] implemented linked brushing by default. Likewise, Mondrian [@theus2002] 

## Current Options for Interactive Visualizations

In the present day, there are many options for interactive data visualization. Many of them stem from the popular JavaScript visualization libraries such as D3.

I argue that, in these more recent interactive data visualization packages, the conceptual ambiguity about what counts as "interactive visualization" described in Chapter \@ref(what-is-interactivity) has seeped into the implementation. For example, the [R Graph Gallery entry on Interactive Charts](https://r-graph-gallery.com/interactive-charts.html) [@holtz2022interactive] several visualizations which allow for surface-level user interactions such zooming, panning, hovering, 3D rotation, and node repositioning within a network graph. However, in all of the examples, the user only affect graphical attributes of a single plot, and so these visualizations do not meet the linked and parametric definitions of interactivity. In contrast, the [Plotly Dash documentation page on Interactive Visualizations](https://dash.plotly.com/interactive-graphing) [@bibentry2022plotly] does feature two examples of linked hovering and cross-filtering, i.e. examples of linked interactivity. However, it should be noted that vast majority of visualizations in the [Plotly R Open Source Graphing Library documentation page](https://plotly.com/r/) [@bibentry2022plotly] allow for only surface-level interactions. Similarly, [VegaLite Gallery pages on Interactive Charts](https://vega.github.io/vega-lite/examples/#interactive-charts) and [Interactive Multiview Displays](https://vega.github.io/vega-lite/examples/#interactive-multi-view-displays)  [@bibentry2022vegalite] feature many examples, however, only a few show limited examples of linked or parametric interactivity.  Finally, the [Highcharter Showcase Page ](https://jkunst.com/highcharter/articles/showcase.html) [@kunst2022] does not feature any examples of linking or parametric interactivity.

What all of the packages listed above have in common is that most interaction is typically surface-level and takes place within a single plot, and the examples that feature interesting types of interactivity (linked or parametric) that are present require an extensive setup (i.e. many lines of code). The main reason for this is most likely that all of these packages have been designed to be very general-purpose and flexible, and the price to pay for this flexibility is that complex types of interactivity require complex code. Another reason is that these packages have been built for static visualizations first, and interactivity second. Further, since all of these packages are native to JavaScript, the expectation may be that if more interesting types of interactivity are desired, the interactive "back-end" may be written separately, outside of the package. Finally, the typical use case for these packages seems to be presentation, not EDA.  

Be it as it may, there is a fairly high barrier for entry for creating interesting types of interactivity (i.e. linked or parametric) with these packages. This may not be an issue for large organizations which can afford to hire specialists to work on complex interactive dashboards and visualizations. However, to the average applied scientist or data analyst, the upfront cost may be too high, especially if one is only interested in exploratory data analysis for one's own benefit. This may be the reason why interactive visualization is nowadays mainly used for data communication, not data exploration [@batch2017interactive]
   



