# Use

## A minimal example: linked scatterplot and barplot

To get started with `plotscape`, we first need to load the package:

```{r}

library(plotscapeTest)

```

Now, the first step of any `plotscape` visualization is creating a visual scene. We save the scene as an object and print it:

```{r}

scene1 <- ps_scene(mtcars) 
scene1

```

Notice that all you can see now is a gray rectangle. This is because we have not populated the scene yet. Let's add a scatterplot and a barplot:

```{r}

scene1 |>
  ps_scatter(ps_map(x = "wt", y = "mpg")) |>
  ps_bar(ps_map(x = "am")) 
```

We use the base R pipe `|>` operator to take an existing `plotscape` scene and append plots to it using plot-wrapper functions with the `ps_{plot type}()` syntax. Instead of the native pipe, we could have also used the `%>%` pipe operator from the `magrittr`/`tidyverse` packages, but that would require us loading an additional package, so let's keep using the native pipe here instead. 

The plot-wrapper functions take in several arguments, the most important of which is the `mapping` argument. This specifies which variables in the data map onto which visual aesthetic or encoding, similar to the `ggplot2` package. We specify the mapping using the `ps_map()` function which takes in named string arguments where the name specifies the aesthetic mapping and the string specifies that name of the variable (column) in the data. The aesthetics can also be more tersely specified by using positional argument matching, without providing explicit names for the aesthetics, like so:

```{r, eval = FALSE}
scene1 |> 
  ps_scatter(ps_map("wt", "mpg")) |>
  ps_bar(ps_map("am")) 
```

Or by supplying a named list:

```{r, eval = FALSE}
scene1 |> 
  ps_scatter(list(x = "wt", y = "mpg")) |>
  ps_bar(list(x = "am")) 
```

(In fact, all the `ps_map()` function do is take the provided arguments and convert it to a named list. One advantage it has over just `list()` is the positional matching, i.e. `ps_scatter(list("wt", "mpg"))` won't work)

But enough about functions and arguments, the hallmark of `plotscape` is interactivity! Try interacting with the barplot by clicking on the right bar:

```{r}
scene1 |> 
  ps_scatter(ps_map(x = "wt", y = "mpg")) |>
  ps_bar(ps_map(x = "am")) 
```

You should now see all of the automatic cars highlighted in the scatterplot. Now, for change, try selecting the four points in the top-left corner of the scatterplot by clicking-and-dragging. To spice things up, hold down the "1" key on your keyboard while dragging assign the points to Group 1 (and color them green!).

```{r}
scene1 |> 
  ps_scatter(ps_map(x = "wt", y = "mpg")) |>
  ps_bar(ps_map(x = "am")) 
```

If you're struggling to see the points, try increasing their size with the "+" key (and if they get too big for your liking, you can make them smaller again with "-"). Or try increasing the transparency of the points with "[" (or decreasing with "]"):

```{r}
scene1 |> 
  ps_scatter(ps_map(x = "wt", y = "mpg")) |>
  ps_bar(ps_map(x = "am")) 
```

There are many different interactive actions you can take. Fortunately, you do not have to remember them all, they're written in the pop-up help bar that you can access by pressing the button marked "?" in the top-right corner of the scene (the grey box around the plots).

You might be getting a bit bored of the old scatterplot and barplot at this point. How about adding a histogram and a squareplot into the mix:

```{r}

ps_scene(mtcars) |>
  ps_scatter(ps_map(x = "wt", y = "mpg")) |>
  ps_bar(ps_map(x = "cyl")) |>
  ps_histo(ps_map(x = "disp")) |>
  ps_wrapper_plot("square", ps_map(x = "gear", y = "am"))

```

Now, in the squareplot, try selecting the square that encodes all 5-gear automatic (hint: the top-right square) and look at the scatterplot. Notice anything interesting?

All of the 5-gear automatic cars are neatly aligned below the main diagonal, meaning that they are less efficient for their weight than we would expect. To arrive at an insight like this with traditional static plots, we would either have to know where to look or try out many different plots (personally, I've used the `mtcars` dataset for years to test models and algorithms and never noticed this trend). With `plotscape`, we can generate insights on the fly!   

## Creating more complex plots

We can get a bit more creative by adding plots programmatically. For example, here's how we can create a "scatterplot matrix"-like visualization of the `iris` data (i.e. one scatterplot per unique combination of two variables):

```{r}

# Create a matrix of all unique combinations  
# of 2 variables in the iris data
vars <- names(iris)
xy <- matrix(vars[combn(1:4, 2)], 
                    ncol = 2, byrow = TRUE)

scene2 <- ps_scene(iris)
for (i in seq_len(nrow(xy))) {
  scene2 <- scene2 |> ps_scatter(ps_map(xy[i, 1], xy[i, 2]))
}
scene2

```

**Note to self: fix sizing, works in regular RMarkdown**


