---
title: "The Fabric of Interactive Visualization"
subtitle: "From the Algebra of Graphics, Statistics, and Interaction to Practical Implementation"
author: "Adam Bartonicek"
site: bookdown::bookdown_site
documentclass: book
classoption: openany
always_allow_html: true
output:
  bookdown::gitbook:
    css: styles.css
  bookdown::pdf_document2:
    latex_engine: xelatex
  bookdown::pdf_book:
    latex_engine: xelatex
    toc: false
    includes:
      before_body: title.tex
toc_depth: 3
bibliography: [references.bib]
biblio-style: "apalike"
link-citations: true
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{tikz-cd}
  - \usepackage{changepage}
  - \AtBeginDocument{\frontmatter}
  - \AtBeginDocument{\let\maketitle\relax}
---

# Abstract

Interactive data visualization has become a staple of modern data presentation. However, creating rich interactive figures still remains challenging, and widely available interactive data visualization systems often offer only limited support for many valuable interactive features. This implementation gap may stem from a subtle yet profound issue: what we can *do* with a figure is fundamentally determined by what the figure *is*. Graphs are not mere images; they are visual representations of structured information. Consequently, interactive features must align with the mathematical properties of the data they represent. Yet, this is view poses challenges for the prevailing, independence-based approach to building data visualization systems. This approach, popularized by the Grammar of Graphics [@wilkinson2012], attempts to treat graphics, statistics, and interaction as independent, modular components.

In this thesis, I argue that the independence model, despite many advantages, is fundamentally inadequate for comprehensively modeling graphics - visualization components are not independent, but instead deeply interconnected, particularly when interaction comes into play. Throughout this work, I explore this interconnectedness, the challenges is poses for interactive data visualization systems, and propose an alternative model based on some fundamental concepts from category theory. To put these ideas into practice, I also present `plotscaper`, an R package developed to refine these theoretical ideas and offer a practical tool for interactive data exploration. My goal is to demonstrate that, to build truly robust and consistent interactive data visualization systems, we may benefit from grounding our thinking in algebraic concepts.

```{r}
#| echo: false
knitr::opts_chunk$set(
  fig.align = "center"
)

directory <- tempdir()

get_ts_filepath <- function(x) {
  if (endsWith(x, ".ts")) {
    paste0(directory, "/", trimws(gsub("//", "", x)))
  } else {
    tempfile(tmpdir = directory, fileext = ".ts")
  }
}

# Create a TypeScript knitr engine
knitr::knit_engines$set(ts = function(options) {
  code <- paste0(options$code, collapse = "\n")
  out <- ""
  
  if (options$eval) {
    ts_path <- get_ts_filepath(options$code[1]) 
    ts_conn <- file(ts_path)
    writeLines(code, ts_conn)
    
    out <- paste("~/.bun/bin/bun run", ts_path) |> 
      system(intern = TRUE, ignore.stderr = TRUE)
    
    close(ts_conn)
  }
  
  knitr::engine_output(options, code, out)
})

```

```{r}
#| echo: false

pal_paired_1 <- palette.colors(2, "Paired")
pal_paired_3 <- palette.colors(6, "Paired")[c(1, 2, 5, 6, 3, 4)]
pal_light_3 <- pal_paired_3[1:3]
pal_dark_3 <- pal_paired_3[2 * 1:3]

library(ggplot2)

clean_theme <- theme_bw() +
  theme(
  axis.ticks = element_blank(),
  panel.grid = element_blank()
)

theme_set(clean_theme)
options(width = 60)

knitr::opts_chunk$set(
  comment = "",
  fig.align = "center",
  out.width = "100%" # For LaTeX output to render correctly
)
```
