---
title: "The Fabric of Interactive Visualization"
subtitle: "From the Algebra of Graphics, Statistics, and Interaction to Practical Implementation"
author: "Adam Bartonicek"
site: bookdown::bookdown_site
documentclass: book
always_allow_html: true
output:
  bookdown::gitbook:
    css: styles.css
  bookdown::pdf_document2:
    latex_engine: xelatex
  bookdown::pdf_book:
    latex_engine: xelatex
bibliography: [references.bib]
biblio-style: "apalike"
link-citations: true
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{tikz-cd}
---

# Abstract

Interactive data visualization has become a staple of modern data presentation. However, creating rich interactive figures still remains challenging. Certain useful interactive features are often missing or only partially supported in current tools. This implementation gap may stem from a subtle yet profound issue: what we can *do* with a visualization is fundamentally determined by what the visualization *is*. Graphs are not mere images; they are visual representations of structured information. Therefore, interactive features must align with the mathematical properties of the data they represent. This seemingly self-evident point clashes with prevailing practice. Many current systems, particularly ones based on the Grammar of Graphics model, treat graphics, statistics, and interaction as independent, modular components that can be freely combined.

In this thesis, I argue that this widely used independence model, despite its benefits, is fundamentally flawed: visualization components are deeply interconnected, not independent. I explore this interconnectedness, the challenges is poses for interactive data visualization systems, and propose a simple, alternative model based on some fundamental concepts from category theory. To put these ideas into practice, I also present `plotscaper`, an R package developed to refine these theoretical ideas and offer a practical tool for interactive data exploration. My goal is to demonstrate that designing truly consistent and general interactive data visualization systems requires grounding our thinking in algebraic concepts.

```{r}
#| echo: false
knitr::opts_chunk$set(
  fig.align = "center"
)

directory <- tempdir()

get_ts_filepath <- function(x) {
  if (endsWith(x, ".ts")) {
    paste0(directory, "/", trimws(gsub("//", "", x)))
  } else {
    tempfile(tmpdir = directory, fileext = ".ts")
  }
}

# Create a TypeScript knitr engine
knitr::knit_engines$set(ts = function(options) {
  code <- paste0(options$code, collapse = "\n")
  out <- ""
  
  if (options$eval) {
    ts_path <- get_ts_filepath(options$code[1]) 
    ts_conn <- file(ts_path)
    writeLines(code, ts_conn)
    
    out <- paste("~/.bun/bin/bun run", ts_path) |> 
      system(intern = TRUE, ignore.stderr = TRUE)
    
    close(ts_conn)
  }
  
  knitr::engine_output(options, code, out)
})

```

```{r}
#| echo: false

pal_paired_1 <- palette.colors(2, "Paired")
pal_paired_3 <- palette.colors(6, "Paired")[c(1, 2, 5, 6, 3, 4)]
pal_light_3 <- pal_paired_3[1:3]
pal_dark_3 <- pal_paired_3[2 * 1:3]

library(ggplot2)

clean_theme <- theme_bw() +
  theme(
  axis.ticks = element_blank(),
  panel.grid = element_blank()
)

theme_set(clean_theme)
options(width = 60)

knitr::opts_chunk$set(
  fig.align = "center",
  out.width = "100%" # For LaTeX output to render correctly
)
```
